---
title: "Cluster Analysis example"
author: "Sergio Vicente"
date: "20-02-2016"
output: 
  html_document:
    fig_width: 6
    fig_height: 5
    fig_caption: true
---

\ 

* This example was produced as an auxiliary study for the course 
["Introdução ao Marketing Analítico", 4th module, by INSPER](https://www.coursera.org/learn/marketing-analitico/home/week/4), Coursera.org.

* It is based on a small data set of **personal spent** ("gastos diários") (clothing and food). It demonstrates some techniques for Cluster Analysis.
   
\ 

#### Reading data

```{r}
dt <- read.table("Gastos_Diarios.csv", sep=',', header=TRUE, row.names=1)
names(dt) <- c("comida","roupas")
print(head(dt))
```

\ 

#### Plotting scattergram (XY graph)

```{r, echo=FALSE}
# http://www.statmethods.net/graphs/scatterplot.html
plot(dt$comida, dt$roupas, main="Comida x Roupas\n(Food x Clothing)", xlab="Comida", ylab="Roupas")
# http://www.statmethods.net/advgraphs/axes.html
text(dt$comida, dt$roupas, row.names(dt), cex=0.6, pos=4, col="blue") 
# print("Graphic XY (comida vs. roupas) is plotted.")
```

Prepare data - standardization
```{r}
dts <- na.omit(dt) # listwise deletion of missing
dts <- scale(dts)  # standardize variables 
print(dts)
```


-------
#### Analysing clusters

Visualizing the Dendrogram
```{r}
# https://rpubs.com/gaston/dendrograms
hc = hclust(dist(dt))   		# prepare hierarchical cluster
plot(hc, xlab="pessoas")		# very simple dendrogram
box(which="figure",lty="solid",col="red",bg="yellow")
```

\ 

Establishing number of clusters solution
```{r}
# http://www.statmethods.net/advstats/cluster.html
wss <- (nrow(dts)-1)*sum(apply(dts,2,var))
# nCenters <- nrow(dts)-1   # (maximum allowed)
nCenters <- 2  			# (chosen visually, observing dendrogram image)	
```

Partioning by K-means
```{r}
for (i in 2:nCenters) 
  wss[i] <- sum(kmeans(dts, centers=i)$withinss)
plot(1:nCenters, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 
```

K-Means Cluster Analysis
```{r}
fit <- kmeans(dts, nCenters) 
# get cluster means
aggregate(dts,by=list(fit$cluster),FUN=mean)
# append cluster assignment
dt.cluster <- data.frame(dts, fit$cluster) 
```

#### Resulted Clusters 
```{r, echo=FALSE}
print(head(dt.cluster))
```

\ 

#### Plotting the reviewed scattergram with clusters determined
```{r, echo=FALSE,fig.height=7,fig.width=8}
# Set up the plot region:
par(bg="#fce8e4")
plot(dt$comida, dt$roupas, main="Comida x Roupas - CLUSTERS\n(Food x Clothing)", sub="- personal spent -", xlab="Comida", ylab="Roupas", col.main="navy", pch=19)
text(dt$comida, dt$roupas, dt.cluster$fit.cluster, cex=1.2, pos=2, col="blue", font=4) 
text(dt$comida, dt$roupas, row.names(dt), cex=1.1, pos=4, col="maroon", font=2) 

# http://stackoverflow.com/questions/22463710/filled-rectangle-in-r
u <- par("usr")
rect(u[1], u[3], u[2], u[4], col = "#fce802", border = "black", density=20, lty=2, angle=90) 
box(which="figure",lty="solid",col="red",bg="yellow")
mtext( "legend:\n  Clusters: 1,2\n  Pessoas: a-e", cex=1, side=1, adj=-0.1, line=3.4 )

# print("XY cluster analysis (comida vs. roupas) is plotted.")
```

\ 

This [R] code is available to see and download from my GitHub:
[https://github.com/svicente99/cluster_analysis_example](https://github.com/svicente99/cluster_analysis_example)

\ 

